Will Kamala Harris will win the 2024 Presidential election

### 2024-11-04 paper/paper.qmd
merged commits

To go a little more in-depth into what we actually did, we have to first discuss our estimand. Our estimand, the parameter we aim to study, is the predicted support for Kamala Harris in the swing states (measured in %). We do this by creating Baysian model with 3 parameters; state, pollster, and says since Kamala Harris was announced as candidate. We have the priors set with a normal distribution centered at 0, with a standard deviation of 2.5. We expect that because the swing states are states that don't have a party that they dominantly vote for, the expected support will be around 50%. We also make other graphs in the Data section, which helps us understand the values themselves.

From the result section we find that the average predicted support for Kamala Harris for the 7 swing states were as follows: Arizona at 46.26%, Nevada at 46.43%, Georgia at 46.59%, North Carolina at 47.26%, Michigan at 47.30%, Pennsylvania at 48.02%, Wisconsin at 48.38%. Based on this we predict that Arizona, and Georgia will vote Donald trump and the rest of the states will vote Kamala Harris.

Predicting who will become president is important because it helps people understand and prepare for future changes in the country’s policies and priorities. A new president can shape laws, affect the economy, influence international relationships, and address key issues like healthcare, education, and climate change. Accurate predictions allow citizens, businesses, and governments to plan and adapt. For example, if polls indicate a particular candidate might win, groups can start thinking about how their plans might align with the candidate's proposed policies. It also helps keep people informed and engaged in the political process, making them more likely to vote and participate in democracy.
Results paragraph

@@

In this analysis, we examine polling trends for Kamala Harris across key swing states, focusing on recent polling data from Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin \[CITE NYT PAPER\]. As shown in @fig-supportstate, which displays the support percentages for both Harris and Donald Trump side-by-side by state, the data reveals a competitive race, with relatively close support levels for both candidates in most states. While Trump has a slight edge in Arizona, Georgia, and North Carolina, Harris maintains a modest lead in Michigan, Nevada, and Pennsylvania.

@@

# Plot the second subset
ggplot(subset2, aes(x = end_date, y = pct, color = state)) +
  geom_point() +
  geom_smooth(se = TRUE) +
  facet_wrap(~ state, scales = "free_y") +
  labs(y = "Harris Percent", x = "Date") +
  theme_minimal() +
      scale_x_date(
    date_breaks = "1 month",
    date_minor_breaks = "1 week",
    date_labels = "%b %d" 
  ) +
   theme(legend.position = "none")
```

@fig-supportAM further examines Harris's support trends in Arizona and Michigan. Both states exhibit an upward trend in support starting in early October, which may suggest increasing favorability or strategic campaign efforts in these regions. This upward trajectory could indicate an opportunity for Harris to solidify her support in these competitive areas.

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{State}_i + \beta_2 \times \mbox{Pollster}_i + \beta_3 \times \mbox{Days since first poll}_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
The outcome variable in this model is the proportion of respondents who support Harris in each poll, represented as $y_i$. This binomial outcome (successes out of the total sample size) is modeled as a function of several predictor variables that account for differences by state, pollster, and time.

@@

In @fig-modelNPW, we see states of Nevada, Pennsylvania and Wisconsin have a singular declining trend.

@@

```{r}

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

# Results

@@

Lastly, if we consider @fig-supportWP we see that harris has been polling at a consistent 48% since mid september which is somewhat similar to Georgia, while Pennsylvania seems to be leaning more democratic than republican than Georgia.\@fig-supportGNN, they have both similarly increased their interest in Harris since the polling began until mid september which is when they both becan to plateau at their current polling numbers.

@@

# Appendix 1

Why we choose NYT? - `numeric_grade` is 3 - `pollscore` is -1.5, a score of reliability called "Predictive Optimization of Latent skill Level in Surveys, Considering Overall Record, Empirically.", where negative numbers are better - `transparency_score` is 9, reflects pollsters transparency about their methodology (calculated based on how much information it discloses about its polls and weighted by recency) - `population_full` is 'rv', respondents are registered voters

# Appendix 2
Here we will be talking about how we would conduct a $100,000 surevey to gather data about this upcoming election to ensure there is minimal error in the data we are collecting and, our data represents the facts we want them to represent. 
Out of the \$100,000 we will allocate \$15000 for the necessary development and administration of the data. This means we will be spending this money to create the survey, and have a infrastructure in place to hold the large amount of data as well as cover any security fees to keep this data private and safe. Then we allocate $50000 to advertise the survey. We will need a large sample so we will be spending most of our budget for this. We will have a different urls for each advertisement we have so that we can incorporate this in our data to see which demographic is accessing the survey through which platform. The platforms to advertise this is through spotify, facebook, instagram, and some news networks. If possible we will try to use this money to get it endorsed by branches of government to show our reliability. Reliability is also what citizens respond to so this will get a lot of respondents. We will spend \$25,000 for modeling our data as they tend to be expensive with data this size. And the rest of the \$15,000 will be leftover cost for anything that we don't foresee. If there is any leftover we can add a survey participation price such as having a free 3 month trial for the platforms we mentioned above (that is if they have a subscription based membership).

We will not be using telephone surveys because according to reserach [@idSurvey] we find the most people, don't pick up calls from unknown numbers and even the people that pick it up they are less inclined to answer the questions of the survey.

@@

Finally we take a look at how we sample from the gathered data. We thought about random sampling but we were worried about having unequal proportions of the demographic we would advertise to. So if more people that go to facebook responded, that doesn't necessarily mean that people use facebook more than the other platforms. So we we decided to combine stratified and cluster  sampling(@annurev). The idea is that we will choose based on the different platforms we advertise to first to have a cluster sampling method. And then we will sample the data by dividing the people by the certain aspects of the surevy, like if they are in different age groups or if they are registered voters. This is called Stratified Sampling (@annurev). Then we take a union of all the samples and we gather a group that we are able to build models based off of.
There are obvious disadvantages as well. For example, we have fewer questions meaning when we sample based on the question we might have cases where, some groups are over or under represented. This can happen for example when most of the people who answered might be from age 18-50 because those are the age groups with access to a phone and are frequently visiting the sites we advertise our survey on. Also the different apps may skew the data as Facebook has a older audience as opposed to Instagram and Tiktok. Another disadvantage is the fact that we have 2 different sampling methods that we use and we combine them both, meaning that this can be an expensive process. Doing so may exceed the allocated budget however, it will be worth it because we minimize the sampling bias we talk about earlier where we may have a group that has more access to the survey. The last disadvantage we also foresee is the fact that younger audience may not be as interested in filling out a survey about politics as they feel, it might not directly affect them. We try to minimize this by offering compensation for their time when they fill out the survey, i.e. free membership services for 3 months on Spotify.

@@

# Appendix 3 - checks and summary stats {#sec-checks}

# Plot checks for Model
pp_check(bay_model)
plot(bay_model, pars = "(Intercept)", prob = 0.95)
modelsummary::modelsummary(
  list(
    "First model" = bay_model
  ),
  statistic = "mad",
  fmt = 2
)
```

>>>>

To go a little more in-depth into what we actually did, we have to first discuss our estimand. Our estimand, the parameter we aim to study, is the predicted support for Kamala Harris in the swing states (measured in %). We do this by creating Baysian model with 3 parameters; state, pollster, and says since Kamala Harris was announced as candidate. We have the priors set with a normal distribution centered at 0, with a standard deviation of 2.5. We expect that because the swing states are states that don't have a party that they dominantly vote for, the expected support will be around 50%. We also make other graphs in the Data section, which helps us understand the values themselves. 
From the result section we find that the average predicted support for Kamala Harris for the 7 swing states were as follows: Arizona at 46.26%, Nevada at 46.43%, Georgia at 46.59%, North Carolina at 47.26%, Michigan at 47.30%, Pennsylvania at 48.02%, Wisconsin at 48.38%. Based on this we predict that Arizona, and Georgia will vote Donald trump and the rest of the states will vote Kamala Harris.

Predicting who will become president is important because it helps people understand and prepare for future changes in the country’s policies and priorities. A new president can shape laws, affect the economy, influence international relationships, and address key issues like healthcare, education, and climate change. Accurate predictions allow citizens, businesses, and governments to plan and adapt. For example, if polls indicate a particular candidate might win, groups can start thinking about how their plans might align with the candidate's proposed policies. It also helps keep people informed and engaged in the political process, making them more likely to vote and participate in democracy.

@@

In this analysis, we examine polling trends for Kamala Harris across key swing states, focusing on recent polling data from Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin [CITE NYT PAPER]. As shown in @fig-supportstate, which displays the support percentages for both Harris and Donald Trump side-by-side by state, the data reveals a competitive race, with relatively close support levels for both candidates in most states. While Trump has a slight edge in Arizona, Georgia, and North Carolina, Harris maintains a modest lead in Michigan, Nevada, and Pennsylvania.

@@

# Plot the second subset
ggplot(subset2, aes(x = end_date, y = pct, color = state)) +
  geom_point() +
  geom_smooth(se = TRUE) +
  facet_wrap(~ state, scales = "free_y") +
  labs(y = "Harris Percent", x = "Date") +
  theme_minimal() +
      scale_x_date(
    date_breaks = "1 month",
    date_minor_breaks = "1 week",
    date_labels = "%b %d" 
  ) +
   theme(legend.position = "none")
```
@fig-supportAM further examines Harris's support trends in Arizona and Michigan. Both states exhibit an upward trend in support starting in early October, which may suggest increasing favorability or strategic campaign efforts in these regions. This upward trajectory could indicate an opportunity for Harris to solidify her support in these competitive areas.

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 \times \mbox{State}_i + \beta_2 \times \mbox{Pollster}_i + \beta_3 \times \mbox{Days since first poll}_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```

The outcome variable in this model is the proportion of respondents who support Harris in each poll, represented as $y_i$. This binomial outcome (successes out of the total sample size) is modeled as a function of several predictor variables that account for differences by state, pollster, and time.

@@

```
In @fig-modelNPW, we see states of Nevada, Pennsylvania and Wisconsin have a singular declining trend.

@@

@@

Lastly, if we consider @fig-supportWP we see that harris has been polling at a consistent 48% since mid september which is somewhat similar to Georgia, while Pennsylvania seems to be leaning more democratic than republican than Georgia.@fig-supportGNN, they have both similarly increased their interest in Harris since the polling began until mid september which is when they both becan to plateau at their current polling numbers.

@@

# Appendix 1 - Pollster Methodology

The New York Times/Siena College Poll is a respected polling partnership that provides data-driven insights into U.S. public opinion, particularly on political, social, and economic issues. Known for its rigorous methodology, this poll focuses on capturing the views of registered and likely voters across the country, with particular emphasis on key battleground states during election cycles. The poll is frequently cited for its accuracy and transparency, and is widely regarded as reliable by the public, political analysts, and the media.

The New York Times/Siena College Poll is conducts live telephone interviews with a national sample of registered voters, using both landlines and cellphones, with over 90% of contacts made via cellphones. The sample frame is drawn from a comprehensive voter file covering 200 million registered voters. This file provides demographic and geographic information, enabling representative sampling across age, race, and political affiliation. The poll targets registered U.S. voters, with special focus on battleground states like Arizona, Georgia, Michigan, Nevada, Pennsylvania, and Wisconsin, determined as swing states. Respondents are randomly selected from the voter file, ensuring a diverse sample by region, party, and demographics. To mitigate non-response bias, the poll applies demographic weighting, increasing the weight of underrepresented groups, such as young adults or non-college graduates. Low response rates (around 2%) present challenges, to mitigate this, weighing adjustments are applied to correct for demographic imbalances and potential non-response bias. Questions are structured to be clear, neutral, and representative across political views, with each question carefully reviewed for balance and clarity. This process reduces the risk of bias in responses, though the brief format of phone surveys limits depth.

# Appendix 2 - Survey

Here we will be talking about how we would conduct a $100,000 surevy to gather data about this upcoming election to ensure there is minimal error in the data we are collecting and, our data represents the facts we want them to represent. 
Out of the \$100,000 we will allocate \$15000 for the necessary development and administration of the data. This means we will be spending this money to create the survey, and have a infrastructure in place to hold the large amount of data as well as cover any security fees to keep this data private and safe. Then we allocate $50000 to advertise the survey. We will need a large sample so we will be spending most of our budget for this. We will have a different urls for each advertisement we have so that we can incorporate this in our data to see which demographic is accessing the survey through which platform. The platforms to advertise this is through spotify, facebook, instagram, and some news networks. If possible we will try to use this money to get it endorsed by branches of government to show our reliability. Reliability is also what citizens respond to so this will get a lot of respondents. We will spend \$25,000 for modeling our data as they tend to be expensive with data this size. And the rest of the \$15,000 will be leftover cost for anything that we don't foresee. If there is any leftover we can add a survey participation price such as having a free 3 month trial for the platforms we mentioned above (that is if they have a subscription based membership).

We will not be using telephone surveys because according to research [@idSurvey] we find the most people, don't pick up calls from unknown numbers and even the people that pick it up they are less inclined to answer the questions of the survey.

@@

Finally we take a look at how we sample from the gathered data. We thought about random sampling but we were worried about having unequal proportions of the demographic we would advertise to. So if more people that go to facebook responded, that doesn't necessarily mean that people use facebook more than the other platforms. So we we decided to combine stratified and cluster  sampling(@annurev). The idea is that we will choose based on the different platforms we advertise to first to have a cluster sampling method. And then we will sample the data by dividing the people by the certain aspects of the surevy, like if they are in different age groups or if they are registered voters. This is called Stratified Sampling (@annurev). Then we take a union of all the samples and we gather a group that we are able to build models based off of.

There are obvious disadvantages as well. For example, we have fewer questions meaning when we sample based on the question we might have cases where, some groups are over or under represented. This can happen for example when most of the people who answered might be from age 18-50 because those are the age groups with access to a phone and are frequently visiting the sites we advertise our survey on. Also the different apps may skew the data as Facebook has a older audience as opposed to Instagram and Tiktok. Another disadvantage is the fact that we have 2 different sampling methods that we use and we combine them both, meaning that this can be an expensive process. Doing so may exceed the allocated budget however, it will be worth it because we minimize the sampling bias we talk about earlier where we may have a group that has more access to the survey. The last disadvantage we also foresee is the fact that younger audience may not be as interested in filling out a survey about politics as they feel, it might not directly affect them. We try to minimize this by offering compensation for their time when they fill out the survey, i.e. free membership services for 3 months on Spotify.

@@

# Appendix 3- Posterior Predictive Check {#sec-checks}

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-ppcheck
#| fig-cap: "Posterior predictive check of model"

# Plot checks for Model
pp_check(bay_model)

In @fig-ppcheck, we check the Posterior predictive check of model of our model. Since the raw data aligns closely with our prdictions, our model is a good predicton of the data.
